Cooperative Kernels: Safe Multitasking for Blocking Algorithms on GPUs

### 2016-11-15 write-ups/PLDI/paper.tex
Fixed undefined references.

Concurrent calls to $\offerfork()$ must be protected with a mutex,
otherwise workgroups calling $\offerfork()$ will conflict over the
available workgroups. The same conflict over available workgroups can
occur at a new kernel launch between the scheduler and a workgroup
calling $\offerfork()$ occurs between the scheduler assigning tasks.
Thus we provide a mutex, called the \emph{assignment mutex}, in the
scheduler context which must be held anytime a group of available
resources is assigned to a kernel.

The provides an indirect mutex protecting concurrent calls to
$\offerfork()$ and $\offerkill()$. That is, $\offerkill()$ only
commands a workgroup to exit when the scheduler holds the
\emph{assignment lock}, as the scheduler is trying to obtain work
groups to send to a task, and $\offerfork()$ cannot spawn workgroups
without holding this lock.

The semantics of $\offerkill()$, which allow only the workgroup with
the highest id provide natural mutual exclusion for exiting workgroups
(as part of the exit, the kernel context is updated). That is, a
workgroup can only exit at $\offerkill()$ if it is the workgroup with
the highest id. An exiting thread releases this natural mutex by
decrementing the number of active workgroups in the kernel context.

@@

In section~\ref{??}, we show that a resizing barrier is semantically
equivalent to three normal barrier calls with calls to cfork and ckill
in between. It is possible, however to implement the resizing with
only one call to a global barrier with $\offerfork$ and $\offerkill$
baked into the implementation. \TSComment{note here that these
  implementations require no cfork or ckill outside of the barrier}

We consider barriers which have the master/slave model
(e.g. \cite{??}). That is, one workgroup (the master) collects signals
from the other workgroups (the slaves) indicating that they have
arrived at the barrier and are waiting for a reply indicating that
they may leave the barrier. Once the master has recieved a signal from
all the slaves, it replies with a signal saying that the slaves may
leave. \TSComment{Need to discuss new active workgroup number that
is stable across iterations}

@@

We observe if we provide an additional scheduler functionality for a
workgroup executing a cooperative kernel, we can greatly reduce the
gather time. We call the new function {\tt query()}. The semantics
simply return the number of workgroups that the scheduler requires to
kill from the cooperative kernel. A new barrier can now be implented
as follows: (1) Master waits for all slaves; (2) Master forks and sets
new workgroups as slaves. (3) Master calls {\tt query()} and gets
value $W$; (4) master releases slaves and broadcasts the value from
{\tt query} to slaves. (5) Workgroups with id greater than $M-W$ spin
on $\offerkill$, as they know from {\tt query()} that the scheduler
needs them to exit. We show in Section~\ref{} that the barrier using
query greatly reduces the gather time in practice.

\subsection{Vendor Implementation Defects}

\ADComment{Briefly rant about our pain.  Compiler hang, compiler crashes, deadlocks, defunct processes.  Caution: don't make it sound so terrible that you can't trust our results.  Gently acknowledge that our implementation may be imperfect.}

@@

Despite its apparent complexity, the $\textsc{Resizing-Barrier}$ rule simply implements the rewriting of $\resizingglobalbarrier$ in terms of $\globalbarrier$, $\offerkill$ and $\offerfork$ discussed in \mysec\ref{ }.

>>>>

Concurrent calls to $\offerfork$ must be protected by a mutex
to avoid racing on the
available workgroups. The same conflict over available workgroups can
occur at a new kernel launch between the scheduler and a workgroup
calling $\offerfork()$.

The scheduler context provides an \emph{assignment mutes} to protecting concurrent calls to
$\offerfork$ and $\offerkill$.  On calling $\offerkill$, a workgroup is only permitted to exit if the scheduler currently holds the assignment mutex; that is, the scheduler is trying to obtain work
groups to send to a task.  Conversely, on calling $\offerfork$ a workgroup can only spawn new workgroups if the assignment mutex is not held by the scheduler.

Concurrent calls to $\offerkill$ are acceptable because only the workgroup with the highest id may exit, automatically ensuring mutual exclusion for exit-related modifications.

@@

In \mysec\ref{sec:semantics}, we defined the semantics of a resizing barrier in terms
of a series of calls to $\globalbarrier$, $\offerfork$ and $\offerkill$.
 It is possible, however to implement the resizing with
only one call to a global barrier with $\offerfork$ and $\offerkill$
baked into the implementation. \TSComment{note here that these
  implementations require no cfork or ckill outside of the barrier}

We consider barriers which have the master/slave model
(e.g.\ \cite{XF10}). That is, one workgroup (the master) collects signals
from the other workgroups (the slaves) indicating that they have
arrived at the barrier and are waiting for a reply indicating that
they may leave the barrier. Once the master has recieved a signal from
all the slaves, it replies with a signal saying that the slaves may
leave. \TSComment{Need to discuss new active workgroup number that
is stable across iterations}

@@

We observe if we provide an additional scheduler functionality for a
workgroup executing a cooperative kernel, we can greatly reduce the
gather time. We call the new function {\tt query()}. The semantics
simply return the number of workgroups that the scheduler requires to
kill from the cooperative kernel. A new barrier can now be implented
as follows: (1) Master waits for all slaves; (2) Master forks and sets
new workgroups as slaves. (3) Master calls {\tt query()} and gets
value $W$; (4) master releases slaves and broadcasts the value from
{\tt query} to slaves. (5) Workgroups with id greater than $M-W$ spin
on $\offerkill$, as they know from {\tt query()} that the scheduler
needs them to exit. We show in \mysec\ref{sec:responsiveness} that the barrier using
query greatly reduces the gather time in practice.

@@

Despite its apparent complexity, the $\textsc{Resizing-Barrier}$ rule simply implements the rewriting of $\resizingglobalbarrier$ in terms of $\globalbarrier$, $\offerkill$ and $\offerfork$ discussed in \mysec\ref{sec:semantics}.
